The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-03 README.Rmd
move helper functions to README, add rockerverse.R
preparing for submission, see #26

also add missing used packages to DESCRIPTION

>>>>

## Helpers

The [README.Rmd](README.Rmd) file's source includes some useful little code snippets, e.g. count words or get a list of all contributor's GitHub handles.

```{r word_stats, eval=FALSE, include=FALSE}
# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
wordcountaddin::readability(here::here("rockerverse.Rmd"))
wordcountaddin::text_stats(here::here("rockerverse.Rmd"))
```

```{r preprint_pdf, eval=FALSE, include=FALSE}
# run this chunk, get the PDF, then revert changes in touched files
library("rmarkdown")
library("rticles")
unlink('manuscript_cache', recursive = TRUE)

# function is a copy of rticles::rjournal_article() but...
# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
# - directly referencing unexported functions from rticles
# - direclty setting the keep_tex option
rjournal_style_preprint <- function (..., citation_package = "natbib") {
    rmarkdown::pandoc_available("2.2", TRUE)
    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL,
        citation_package = citation_package, ...)
    base$pandoc$to <- "latex"
    base$pandoc$ext <- ".tex"
    base$pandoc$keep_tex <- TRUE # added
    base$post_processor <- function(metadata, utf8_input, output_file,
        clean, verbose) {
      filename <- basename(output_file)
      if (filename != (filename2 <- gsub("_", "-", filename))) {
          file.rename(filename, filename2)
          filename <- filename2
      }
      m <- list(filename = xfun::sans_ext(filename))
      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article",
          "includes", "in_header"))
      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble",
          "header-includes")]))) {
          f <- tempfile(fileext = ".tex")
          on.exit(unlink(f), add = TRUE)
          xfun::write_utf8(preamble, f)
          f
      })
      #t <- find_resource("rjournal_article", "RJwrapper.tex")
      t <- here::here("RJwrapper_preprint_template.tex")
      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine,
          clean = clean)
    }
    base$knitr$opts_chunk$comment <- "#>"
    hilight_source <- rticles:::knitr_fun("hilight_source")
    hook_chunk = function(x, options) {
        if (rticles:::output_asis(x, options))
            return(x)
        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
    }
    hook_input <- function(x, options) paste(c("\\begin{Sinput}",
        hilight_source(x, "sweave", options), "\\end{Sinput}",
        ""), collapse = "\n")
    hook_output <- function(x, options) paste("\\begin{Soutput}\n",
        x, "\\end{Soutput}\n", sep = "")
    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks,
        list(chunk = hook_chunk, source = hook_input, output = hook_output,
            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
    base
}

rmarkdown::render("rockerverse.Rmd", output_format = rjournal_style_preprint())
file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival

# For submission to arXiv:
# 1. Copy the following files into a new folder:
# - figures/*
# - rockerverse.tex
# - RJournal.sty
# - RJreferences.bib
# - RJwrapper.tex
# 2. Run pdfLaTeX directly on RJwrapper.tex (works with RStudio)
# 3. Zip the following files and submit to arXiv:
# - /figures
# - rockerverse.tex
# - RJournal.sty
# - RJwrapper.bbl
# - RJwrapper.tex
```

```{r collab_emails, eval=FALSE, include=FALSE}
front_matter <- rmarkdown::yaml_front_matter("rockerverse.Rmd")
emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
cat(emails, sep = "\n")
```

```{bash collab_handles, eval=FALSE, include=FALSE}
cat rockerverse.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
```

### 2020-03-03 rockerverse.Rmd
move helper functions to README, add rockerverse.R
preparing for submission, see #26

also add missing used packages to DESCRIPTION

```{r clientfunctionalities, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
library("tibble")
library("tidyr")
library("dplyr")

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

@@

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.
```{r dockermachine, eval=FALSE, include=FALSE}
library("dockermachine")
dockermachine::machine_create(driver = "amazonec2")
# machine started: https://console.aws.amazon.com/ec2/v2/home?region=us-east-1#Instances:sort=instanceId
# get config on command line: $ docker-machine env machine
# activate: $ $(docker-machine env test) does not work, because of "
# run container: docker run -it -p 80:80 nginx
# allow inbound HTTP traffic
# http://ec2-3-85-85-61.compute-1.amazonaws.com/ shows "Welcome to nginx"!
# Nice!
# > dockermachine::machine_rm()
# Instance shut down.
```

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remainder of this section presents example Docker images and image stacks featuring R intended for data science.

@@

<!--As a further example, -->

@@

```{bash kaggle_images, eval=FALSE, include=FALSE}
docker pull gcr.io/kaggle-images/rcran
docker pull kaggle/rcran # over 22 months old
docker pull gcr.io/kaggle-images/rstats
docker images | grep kaggle
```

```{bash radiant_images, eval=FALSE, include=FALSE}
docker pull vnijs/rsm-msba-spark
docker images | grep vnijs
```

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accommodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

\label{pipelines}
Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
Containers allow for building complex processing pipelines that are independent of the host programming language.
Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for how to use environment variables, volume mounts, and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
Some packages, e.g., \pkg{containerit}, provide Docker images that can be used very similar to a CLI, but this usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow without thee need for users to directly interact with containers \citep{bennett_outsider_2020}.
Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
Therefore, \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
Using only R functions, an end user can install a module with the \pkg{outsider} package and then call module code to seamlessly integrate a tool into their own R-based workflow.
The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.
<!-- informative examples of how modules work, just as a reminder for @nuest
https://github.com/DomBennett/om..astral/blob/master/examples/astral.R
https://github.com/DomBennett/om..raxml#partitioned-dna-analysis
-->

@@

The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
But the applications also point to the challenges of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
<!-- Tyler W./DC from Gigantum: ideas about ease-of-DIY together with difficulty of composition; shared starting points of Rocker -->
This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
Instructions can be outsourced into distributable scripts and then copied into the image during build, but that makes `Dockerfile`s harder to read.
Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
Despite the different image stacks presented here, the pervasiveness of Rocker images can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker Project's suite of images, more systems will be built successfully, but they may also be built behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

@@

```{r word_stats, eval=FALSE, include=FALSE}
# devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)
wordcountaddin::readability(here::here("manuscript.Rmd"))
wordcountaddin::text_stats(here::here("manuscript.Rmd"))
```

```{r preprint_pdf, eval=FALSE, include=FALSE}
# run this chunk, get the PDF, then revert changes in touched files
library("rmarkdown")
library("rticles")
unlink('manuscript_cache', recursive = TRUE)

# function is a copy of rticles::rjournal_article() but...
# - using RJWrapper_preprint_template.tex (which includes header.tex commands)
# - directly referencing unexported functions from rticles
# - direclty setting the keep_tex option
rjournal_style_preprint <- function (..., citation_package = "natbib") {
    rmarkdown::pandoc_available("2.2", TRUE)
    base <- rticles:::pdf_document_format("rjournal_article", highlight = NULL,
        citation_package = citation_package, ...)
    base$pandoc$to <- "latex"
    base$pandoc$ext <- ".tex"
    base$pandoc$keep_tex <- TRUE # added
    base$post_processor <- function(metadata, utf8_input, output_file,
        clean, verbose) {
      filename <- basename(output_file)
      if (filename != (filename2 <- gsub("_", "-", filename))) {
          file.rename(filename, filename2)
          filename <- filename2
      }
      m <- list(filename = xfun::sans_ext(filename))
      h <- rticles:::get_list_element(metadata, c("output", "rticles::rjournal_article",
          "includes", "in_header"))
      h <- c(h, if (length(preamble <- unlist(metadata[c("preamble",
          "header-includes")]))) {
          f <- tempfile(fileext = ".tex")
          on.exit(unlink(f), add = TRUE)
          xfun::write_utf8(preamble, f)
          f
      })
      #t <- find_resource("rjournal_article", "RJwrapper.tex")
      t <- here::here("RJwrapper_preprint_template.tex")
      rticles:::template_pandoc(m, t, "RJwrapper.tex", h, verbose)
      tinytex::latexmk("RJwrapper.tex", base$pandoc$latex_engine,
          clean = clean)
    }
    base$knitr$opts_chunk$comment <- "#>"
    hilight_source <- rticles:::knitr_fun("hilight_source")
    hook_chunk = function(x, options) {
        if (rticles:::output_asis(x, options))
            return(x)
        paste0("```{=latex}\n\\begin{Schunk}\n", x, "\\end{Schunk}\n```")
    }
    hook_input <- function(x, options) paste(c("\\begin{Sinput}",
        hilight_source(x, "sweave", options), "\\end{Sinput}",
        ""), collapse = "\n")
    hook_output <- function(x, options) paste("\\begin{Soutput}\n",
        x, "\\end{Soutput}\n", sep = "")
    base$knitr$knit_hooks <- rticles:::merge_list(base$knitr$knit_hooks,
        list(chunk = hook_chunk, source = hook_input, output = hook_output,
            message = hook_output, warning = hook_output, plot = knitr::hook_plot_tex))
    base
}

rmarkdown::render("manuscript.Rmd", output_format = rjournal_style_preprint())
file.rename("RJwrapper.pdf", "rockerverse-preprint.pdf") # file for own archival

# For submission to arXiv:
# 1. Copy the following files into a new folder:
# - figures/*
# - manuscript.tex
# - RJournal.sty
# - RJreferences.bib
# - RJwrapper.tex
# 2. Run pdfLaTeX directly on RJwrapper.tex (works with RStudio)
# 3. Zip the following files and submit to arXiv:
# - /figures
# - manuscript.tex
# - RJournal.sty
# - RJwrapper.bbl
# - RJwrapper.tex
```

```{r collab_emails, eval=FALSE, include=FALSE}
front_matter <- rmarkdown::yaml_front_matter("manuscript.Rmd")
emails <- sapply(X = front_matter$author, FUN = function(a) { a$email })
cat(emails, sep = "\n")
```

```{bash collab_handles, eval=FALSE, include=FALSE}
cat manuscript.Rmd | grep '    # https://github.com/' | sed 's|    # https://github.com/|@|'
```

@@

>>>>

```{r clients_table, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
library("tibble")
library("tidyr")
library("dplyr")

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package \citep{cran_sys} to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.

<!-- TRY OUT harbor
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
-->

@@

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.

<!-- TRY OUT dockermachine
library("dockermachine")
dockermachine::machine_create(driver = "amazonec2")
# machine started: https://console.aws.amazon.com/ec2/v2/home?region=us-east-1#Instances:sort=instanceId
# get config on command line: $ docker-machine env machine
# activate: $ $(docker-machine env test) does not work, because of "
# run container: docker run -it -p 80:80 nginx
# allow inbound HTTP traffic
# http://ec2-3-85-85-61.compute-1.amazonaws.com/ shows "Welcome to nginx"!
# Nice!
# > dockermachine::machine_rm()
# Instance shut down.
-->

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
The remainder of this section presents example Docker images and image stacks featuring R intended for data science.

@@

@@

<!-- CHECK IMAGES SIZES
docker pull gcr.io/kaggle-images/rcran
docker pull kaggle/rcran # over 22 months old
docker pull gcr.io/kaggle-images/rstats
docker images | grep kaggle

docker pull vnijs/rsm-msba-spark
docker images | grep vnijs
-->

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accommodate users to be able to read and write hundreds of different spatial raster and vector formats \citep{pebesma_simple_2018,cran_rgdal}.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.

<!-- GET THE NUMBER OF REVERSE IMPORTS AND DEPS FOR rgdal AND sf
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
-->

@@

\label{pipelines}
Second, containers are perfectly suited for **packaging and executing software pipelines** and required data.
Containers allow for building complex processing pipelines that are independent of the host programming language.
Due to its original use case (see&nbsp;\nameref{introduction}), Docker has no standard mechanisms for chaining containers together; it lacks definitions and protocols for how to use environment variables, volume mounts, and/or ports that could enable transfer of input (parameters and data) and output (results) to and from containers.
Some packages, e.g., \pkg{containerit}, provide Docker images that can be used very similar to a CLI, but this usage is cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.
**\pkg{outsider}** ([https://docs.ropensci.org/outsider/](https://docs.ropensci.org/outsider/)) tackles the problem of integrating external programs into an R workflow without thee need for users to directly interact with containers \citep{bennett_outsider_2020}.
Installation and usage of external programs can be difficult, convoluted and even impossible if the platform is incompatible.
Therefore, \pkg{outsider} uses the platform-independent Docker images to encapsulate processes in _outsider modules_.
Each outsider module has a `Dockerfile` and an R package with functions for interacting with the encapsulated tool.
Using only R functions, an end user can install a module with the \pkg{outsider} package and then call module code to seamlessly integrate a tool into their own R-based workflow.
The \pkg{outsider} package and module manage the containers and handle the transmission of arguments and the transfer of files to and from a container.
These functionalities also allow a user to launch module code on a remote machine via SSH, expanding the potential computational scale.
Outsider modules can be hosted code-sharing services, e.g., on GitHub, and \pkg{outsider} contains discovery functions for them.

@@

The R packages and use cases presented above show the growing number of users, developers, and real-world applications in the community and the resulting innovations.
But the applications also point to the challenges of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
This ease-of-DIY in combination with the difficulty of reusing parts from or composing multiple `Dockerfile`s is a further reason for **fragmentation**.
Instructions can be outsourced into distributable scripts and then copied into the image during build, but that makes `Dockerfile`s harder to read.
Scripts added to a `Dockerfile` also add a layer of complexity and increase the risk of incomplete recipes.
Despite the different image stacks presented here, the pervasiveness of Rocker images can be traced back to its maintainers and the user community valuing collaboration and shared starting points over impulses to create individual solutions.
Aside from that, fragmentation may not be a bad sign but may instead be a reflection of a growing market that is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker Project's suite of images, more systems will be built successfully, but they may also be built behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

@@

@@

```{r purl_code, eval=TRUE, include=FALSE}
# generate the reproducible R script every time the R Markdown document is compiled
library("callr")
callr::r_safe(function() { knitr::purl("rockerverse.Rmd") })
```
